Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-16 linear_adia.tex
more adiabatic discussion, plus a new fig of growth rates

\section{Adiabatic disks} 
We now examine the VSI in adiabatic disks ($t_c\to\infty$). For 
analytical discussion we consider the initially isothermal case with 
$\Gamma=1$. We eliminate variables in favor of $\delta v_z$ and
$\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
&\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
  v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
  - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
& -\sigma^2\delta v_z = \gamma c_s^2 \frac{d\Delta}{dz} +
 c_s^2\frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
 \left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
{\bf note: incompressible mode, set delta to zero. only works for
  constant gravity.}

@@

Note that the term with $\ln\rho^{\prime 2}$ in the integrand (third
term on the RHS) represents stabilization by the vertical entropy
gradient when there is no vertical shear. It gives an imaginary
contribution to $\hat{\sigma}$ when $q\neq0$, but this is a factor
$\hat{k}$ smaller than the real (stabilizing) contribution, since we
consider $\hat{k}\gg1$. Thus, instability is expected to be due to the
last term on the RHS of Eq. \ref{adia_integral}.

\subsection{Explicit solutions in the thin-disk limit}
To simplify further, we consider the problem in the thin-disk limit so
that $\ln \rho \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes 
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z,
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. We require
the vertical kinetic energy density $\rho|\delta v_z|^2$ to remain
finite.  Then assuming
$g(\zhat)$ is a polynomial, we require 
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}

Inserting the ansatz Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin}
gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
      \alpha\right)g
    +\underbrace{\left(\alpha^2 - \alpha A - C\right)}_\text{set to zero}\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$ vanish. That
is,
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).
\end{align} 
We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).\label{adia_disp}
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}. 
\end{align}
%Recalling that $B$ depends $\hat{\sigma}$, Eq. \ref{adia_disp} gives
%the eigenfrequency as a function of disk paramters.

% \subsection{Special case of a thin-disk with $\gamma=2$} 
% In principle, we can solve Eq. \ref{adia_disp} for the mode
% frequencies and growth rates. This is complicated in the
% general case, but simplifies considerably for $\gamma=2$.

>>>>

\subsection{Adiabatic perturbations} 
We now extend the above discussion to vertically isothermal disks
($\Gamma=1$) with $\gamma\neq1$ and $t_c\to\infty$, i.e. adiabatic
evolution. We eliminate variables in favor of $\delta v_z$ and
$\Delta \equiv \nabla\cdot\delta\bm{v}$ to obtain
\begin{align}
&\Delta\left(1 + \frac{\gamma c_s^2 k_x^2}{D}\right) = \frac{d\delta
  v_z}{dz} + \frac{\ii k_x^2}{D}\left(\ii c_s^2 \frac{d\ln\rho}{dz}
  - \frac{r}{k_x}\frac{d\Omega^2}{dz}\right)\delta v_z,\label{adia_iso1}\\
& -\sigma^2\delta v_z = \gamma c_s^2 \frac{d\Delta}{dz} +
c_s^2\frac{d}{dz}\left(\frac{d\ln\rho}{dz}\delta v_z\right) +
\left(\gamma-1\right)\frac{d\ln\rho}{dz} \Delta.\label{adia_iso2}
\end{align}
{\bf note: incompressible mode, set delta to zero. only works for
  constant gravity.}

@@

Note that the term with $\ln\rho^{\prime 2}$ in the integrand (third
term on the RHS) represents stabilization by the vertical entropy
gradient when there is no vertical shear. It gives an imaginary
contribution to $\hat{\sigma}$ when $q\neq0$, but this is a factor
$|\epsilon/\hat{k}|$ smaller than the real (stabilizing) contribution,
since $\epsilon\ll 1$ for a thin disk and we (usually) consider
$\hat{k}\gg1$. Thus, instability is expected to be due to the 
last term on the RHS of Eq. \ref{adia_integral}.

\subsubsection{Explicit solutions in the thin-disk limit}
To simplify further, we consider the problem in the thin-disk limit so
that $\ln \rho \simeq -\zhat^2/2$. Eq. \ref{adia_iso3} becomes 
\begin{align}
  0 = \delta v_z^{\prime\prime} - \zhat A \delta v_z^\prime + \left(B
    - C \zhat^2\right)\delta v_z,
\end{align}
with
\begin{align}
  &A \equiv 1 + \ii \epsilon q \hat{k},\\
  &B \equiv \hat{\sigma}^2\left(\frac{1}{\gamma} + \hat{k}^2\right) -
  \left(\frac{1}{\gamma} + \ii \hat{k} \epsilon q\right)\\
  &C \equiv \frac{\left(\gamma-1\right)}{\gamma}\hat{k}^2\left(1 - \frac{\ii
      \epsilon q}{\hat{k}}\right).\label{adia_thin}
\end{align}
Let
\begin{align}
  \delta v_z(\zhat) =
  g(\zhat)\exp\left(\frac{\alpha\zhat^2}{2}\right), \label{adia_ansatz}
\end{align}
where $\alpha$ is a constant to be chosen for convenience. We require
the vertical kinetic energy density $\rho|\delta v_z|^2$ to remain
finite.  Then assuming
$g(\zhat)$ is a polynomial, we require 
\begin{align}
  \real\alpha < \frac{1}{2}. 
\end{align}

Inserting the ansatz Eq. \ref{adia_ansatz} into Eq. \ref{adia_thin}
gives
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime + \left(B +
      \alpha\right)g
    +\underbrace{\left(\alpha^2 - \alpha A - C\right)}_\text{set to zero}\zhat^2 g.
\end{align}
We choose $\alpha$ to make the coefficient of $\zhat^2g$ vanish. That
is,
\begin{align}
  \alpha = \frac{1}{2}\left(A - \sqrt{A^2 + 4C}\right).
\end{align} 
We now have an equation in the same form as Eq. \ref{iso_ode3}:
\begin{align}
  0 = g^{\prime\prime} - \hat{z}\left(A - 2\alpha\right)g^\prime +
  \left(B + \alpha\right)g.
\end{align}
We seek polynomial solutions 
\begin{align}
  g(\zhat) = \sum_{m=0}^M b_m \zhat^m,
\end{align}
% We again seek power-series solutions 
% \begin{align}
%   g(\zhat) = \sum_{m=0}^\infty b_m \zhat^m,
% \end{align}
% which requires
% \begin{align}
%   (m+2)(m+1)b_{m+2} + \left[\left(2\alpha - A\right)m + B + \alpha\right]=0.
% \end{align}
which requires
\begin{align}
  B + \alpha = M\left(A-2\alpha\right).
\end{align}
Inserting the definition of $B$ we get
\begin{align}
  &\hat{\sigma}^2\left(\frac{1}{\gamma}+\hat{k}^2\right) \notag\\ &=
  \left(\frac{1}{\gamma}-\frac{1}{2}\right) +\frac{\ii}{2}\epsilon q
  \hat{k} + \frac{1}{2}\left(2M+1\right)\left(A^2 + 4C\right)^{1/2}\label{adia_disp}. 
\end{align}
%Recalling that $B$ depends $\hat{\sigma}$, Eq. \ref{adia_disp} gives
%the eigenfrequency as a function of disk paramters.

\subsubsection{The case of $\gamma=2$}
In principle, we can solve Eq. \ref{adia_disp} for the mode
frequencies and growth rates. This is complicated in 
general, but simplifies for $\gamma=2$, for which the quantity
$A^2+4C$ is real. In this case we find
\begin{align}
  \hat{\nu}^2 =
  \frac{1}{2}\frac{(2M+1)}{\sqrt{1+2\hat{k}^2}}&\left[
    \sqrt{1-\frac{4M(M+1)\hat{k}^2\epsilon^2q^2}{(2M+1)^2(1+2\hat{k}^2)}}\right.\notag\\
  &\phantom{=}
  \left.- \sqrt{1 - \frac{\hat{k}^2\epsilon^2q^2}{(1+2\hat{k}^2)}}\right].
\end{align}
In the limit of a thin disk $\epsilon\to0$ or weak shear $|q|\to0$, we
have
\begin{align}
  \hat{\nu}^2 = \frac{\epsilon^2
    q^2}{4(2M+1)}\frac{\hat{k}^2}{(1+2\hat{k}^2)^{3/2}},\label{gam2_growth}
\end{align}
and the maximum growth rate occurs at $\hat{k}=\hat{k}_\mathrm{opt}=1$,  
\begin{align}
  \mathrm{max}(\hat{\nu}) &= \frac{|\epsilon
    q|}{2(3^{3/4}\sqrt{2M+1})}\label{gam2_max_growth}.
  % & \leq \frac{|\epsilon
  % q|}{2\times3^{3/4}}\notag.
\end{align}
Eq. \ref{gam2_max_growth} show that growth rates decrease with $M$,
and there exists a maximum possible growth rate at $M=0$. This is
despite the fact that in the thin-disk approximation, for which
$d\Omega^2/dz\propto z$, there is infinite vertical shear
available.

\subsubsection{Comparison with $\gamma=1$}
We can compare growth rates obtained for $\gamma=2$ to those for
$\gamma=1$ in the previous section. If we compare growth rates at their 
respective optimum radial wavelengths, we find
that for $\gamma=2$ the \emph{maximum} possible growth rate, occuring
at $M=0$, is still $\sim2$ times smaller than the \emph{minimum} growth
rate for $\gamma=1$. That is,
\begin{align}
  &\mathrm{max}(\hat{\nu}) \geq \frac{|\epsilon q|}{2\sqrt{1+|\epsilon
      q|}} & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for }   \gamma=1,  \\
  &\mathrm{max}(\hat{\nu}) \leq \frac{|\epsilon q|}{2\times 3^{3/4}}
           & \text{at }\hat{k} = \hat{k}_\mathrm{opt} \text{ for } \gamma=2.
\end{align}

We can also compare growth rates in the limit $\hat{k}\gg1$ using
Eq. \ref{simple_growth} and Eq. \ref{gam2_growth},
\begin{align}
  &\hat{\nu} \geq \sqrt{\frac{|\epsilon q|}{2\hat{k}}}
  &  \text{as }  \hat{k}\to\infty  \text{ for }  \gamma=1, \\
  &\hat{\nu} \leq \frac{|\epsilon q|}{2^{7/4}\sqrt{\hat{k}}}
  &\text{as }  \hat{k}\to\infty  \text{ for }  \gamma=2.
\end{align}
Then the growth rate of perturbations with large $|\hat{k}|$ in a
$\gamma=2$ disk is at most $2^{-5/4}\sqrt{|\epsilon q|}$ times that in a
$\gamma=1$ disk. For $\epsilon = 0.1$ and $|q|=1$, this factor is
$\sim 0.1$.

In Fig. \ref{adia_growth} we plot the growth rate $|\hat{\nu}|$ as a
function of the radial wavenumber $\hat{k}$, for a range of adiabatic
indices, using Eq. \ref{adia_disp}. The presence of a positive
vertical entropy gradient is strongly stabilizing. This is because the
vertical shear grows as $z$ away from the midplane (and has a maximum
when the thin-disk approximation is relaxed), but the
square of the bouyancy frequency, which is stabilizing, grows as $z^2$
away from the midplane. The entropy $P/\rho^\gamma$ formally diverges
for $z\to\infty$ even without the thin-disk approximation.

\begin{figure}
  \includegraphics[width=\linewidth]{figures/rate_theory_gmma}
  \caption{Growth rate of the fundamental VSI mode in vertically
    isothermal disks ($\Gamma=1$) subject to adiabatic perturbations,
    calculated in thin disk approximation from Eq. \ref{adia_disp}
    with $M=0$, $\epsilon=0.1$ and  
    $|q|=1$. Growth rates are shown as a function of the radial
    wavenumber for different values of the adiabatic index
    $\gamma$. For $\gamma\equiv 1$ the perturbations are isothermal
    and there is no vertical entropy gradient.\label{adia_growth}}  
\end{figure}
